The Algorand Virtual Machine

### 2021-03-04 tex/avm.tex
Explain total stake calculation and update the makefile
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

% Preamble
\documentclass[11pt, A4]{article}

@@

\title{The Algorand Virtual Machine Implementation}
\author{aybehrouz}
\date{January 2021}

@@

@@

    To mitigate this problem, for calculating a user's stake at the round \(n\), instead of using the raw ALGO
    balance, we use the minimum of a \emph{trust value} the system has calculated for the user and the user's
    ALGO balance:
    \[
        Stake_{user,n} = \min (Balance_{user,n}, Trust_{user,n})
    \]
    For estimating the value of \(Trust_{user,n}\) we use the following formula:
    \[
        Trust_{user,n} = \max (M_n, \beta Balance_{user,n})
    \]
    Where \(M_n\) is the exponential moving average of the user's ALGO balance at the round \(n\) and \(\beta\) is a
    constant between \(0\) and \(1\) determining the initial trust value of new users.

    In our system a user who held ALGOs and participated in the consensus for a long time is more trusted
    than a new user with a higher balance. An attacker who has obtained a large amount of ALGOs, also needs to hold
    them for a long period of time before being able to attack our system.

@@

    A smart contract oracle is a full AVM emulator that keeps a full local copy of AVM memory and can emulate AVM
    execution without accessing aZK-EDB. Smart contract oracles can be used for reporting useful information about
    \texttt{avmCall} transactions such as accessed AVM heap pages or code area blocks, exact amount of execution cost,
    and soon.

>>>>

% Preamble
\documentclass[11pt, A4]{report}

@@

\title{The Algorand Virtual Machine}
\author{aybehrouz}
\date{January 2021}

@@

    \chapter{Specification}\label{ch:specification}

    \section{Introduction}\label{sec:introduction}

    The Algorand Virtual Machine (AVM) is an abstract computing machine for executing Algorand's smart contracts. It
    is designed in a way that it could be efficiently implemented in either hardware or software.

    \section{The Structure of the Algorand Virtual Machine}\label{sec:the-structure-of-the-algorand-virtual-machine}

    ...

    \subsection{Data Types}\label{subsec:data-types}

    The Algorand Virtual Machine expects that all type checking is done prior to run time, typically by a compiler,
    and does not have to be done by the Algorand Virtual Machine itself.

    The instruction set of the Algorand Virtual Machine distinguishes its operand types using instructions intended to
    operate on values of specific types. For instance, \texttt{iadd} assumes that its operands are two 64-bit integers.

    \subsection{The \texttt{PC} Register}\label{subsec:the-pc-register}

    ...

    \subsection{Call Stack}\label{subsec:call-stack}

    A call stack contains the information that is needed for restoring the state of the invoker of a method.

    \subsection{Memory Unit}\label{subsec:memory-unit}

    The Algorand Virtual Machine has a byte addressable memory which is divided into separate segments. Every segment
    belongs to a smart contract that has a unique \texttt{applicationID}. The AVM always has a single working memory
    segment and memory locations outside its current working segment can not be accessed. The only instructions which
    can change the current working segment are \texttt{invoke\_external}, \texttt{athrow} and return instructions.

    Every smart contract has its own memory segment. Hence, there is no way for a smart contract to access another
    smart contract's memory. Interaction between smart contracts is done using \texttt{invoke\_external} instruction,
    and A smart contract can invoke methods of another smart contract by this instruction.

    A memory segment consists of five data areas:
    \begin{itemize}
        \item Code Area
        \item Constant Area
        \item Local Frame
        \item Operand Stack
        \item Heap
    \end{itemize}
    All data areas except operand stack, have their own address space which starts from 0. Operand stack is a
    last-in-first-out (LIFO) stack and is not addressable. Every memory access instruction operates on its specific
    data areas.

    \subsubsection{Code Area}

    The code area of a segment contains the byte-code of the smart contract which owns that segment. The AVM has no
    instructions for manipulating the code area. \textbf{Installing, removing and updating smart contracts need to
    be done externally.}

    \subsubsection{Constant Area}

    The constant area of a segment contains several kinds of constants, ranging from user defined constants to method
    address tables. A method address table stores method locations in the code area and their access type. The access
    type of a method can be either \texttt{public} or \texttt{private}. The AVM has no instructions for modifying the
    constant area.

    \emph{Only public methods can be invoked by \texttt{invoke\_external} instruction.}

    \subsubsection{Local Frame}

    A local frame is used to store methods parameters and local variables. A new frame is created each time a method
    is invoked. A frame is destroyed when its method invocation completes, whether the completion is normal or abrupt.

    \subsubsection{Operand Stack}

    Every time a local frame is created, a corresponding empty last-in-first-out (LIFO) stack is created too. AVM
    instructions take operands from the operand stack, operate on them, and push the result back onto the operand
    stack. An operand stack is destroyed when its owner method completes, whether that completion is normal or abrupt.

    \subsubsection{Heap}

    The heap of a segment is a persistent memory area which is divided into pages. Each page can be referenced by an
    index that starts from 0. Memory locations inside every page have a separate address space that starts from 0. In
    other words, the address of every memory location inside a heap is a pair of indices: \texttt{(pageIndex, offset)
    }. Pages of a heap do not need to be equally sized.

    \emph{The reason behind this paged design is that the heap is usually persisted using a block device. A heap with
    a paged structure could expose the underlying block based nature of the persistence layer to the application
    layer. In this way, the compiler or the programmer could better optimize the code for the persistence layer.}

    \section{Instruction Set Summary}\label{subsec:instruction-set-summary}

    An Algorand Virtual Machine instruction consists of a \textbf{one-byte} opcode specifying the operation to be
    performed, followed by zero or more operands supplying arguments or data that are used by the operation.
    \textbf{The number and size of the operands are determined solely by the opcode.}

    \subsection{Method Invocation}\label{subsec:method-invocation}

    The Algorand Virtual Machine has three types of method invocation:
    \begin{itemize}
        \item \texttt{invoke\_internal} invokes a method from the current running smart contract.
        \item \texttt{invoke\_external} invokes a \texttt{public} method from another smart contract. It will change
        the current memory segment to the segment of the invoked smart contract.
        \item \texttt{invoke\_native} invokes a method that is not hosted by the Algorand virtual machine. By this
        instruction, high performance native methods of a hosting machine could become available to AVM smart contracts.
    \end{itemize}

    \emph{In the future, we may need to add special instructions for invoking interface and virtual methods...}

    Each time a method is invoked a new local frame and operand stack is created. The Algorand Virtual Machine uses
    local frames to pass parameters on method invocation. On method invocation, any parameters are passed in
    consecutive local variables stored in the method's local frame starting from local variable 0. The invoker of a
    method writes the parameters in the local frame of the invoked method using \texttt{arg} instructions.

    \subsubsection{Exceptions}

    An exception is thrown programmatically using the \texttt{athrow} instruction. Exceptions can also be thrown by
    various Algorand Virtual Machine instructions if they detect an abnormal condition. Some exceptions are not
    catchable and will always abort the execution of the smart contract.

    \subsubsection{Method Invocation Completion}

    A method invocation completes normally if that invocation does not cause an exception to be thrown, either
    directly from the AVM or as a result of executing an explicit throw statement. If the invocation of the current
    method completes normally, then a value may be returned to the invoking method. This occurs when the invoked
    method executes one of the return instructions, the choice of which must be appropriate for the type of the value
    being returned (if any). Execution then continues normally in the invoking method's local frame with the returned
    value (if any) pushed onto the operand stack.

    A method invocation completes abruptly if an exceptions is thrown and is not caught by the current method. A
    method invocation that completes abruptly never returns a value to its invoker.

    When a method completes, whether normally or abruptly, the call stack is used to restore the state of the invoker,
    including its local frame and operand stack, with the \texttt{PC} register appropriately restored and incremented
    to skip past the method invocation instruction. If the invoker was another smart contract, i.e.\ the invocation
    was made by an \texttt{invoke\_external} instruction, the current memory segment will be changed to the invoker's
    segment.

    A thrown exception causes methods in the call stack to complete \textbf{abruptly} one by one, as long as the
    \texttt{PC} register is not pointing to a \texttt{catch} instruction. The \texttt{catch} instruction acts like a
    branch instruction that branches only if an exception is caught. \textbf{When an external method invocation
    completes abruptly, before changing the current segment, all changes made to the heap area will be rolled back.}

    \emph{By using the \texttt{athrow} instruction properly, a programmer can make any method act like an atomic
    operation.}

    \subsubsection{Authorizing Operations}

    In blockchain applications, we usually need to authorize certain operations. For example, for sending an asset
    from a user to another user, first we need to make sure that the sender has authorized this operation. The
    Algorand virtual machine has no built in mechanism for authorizing operations, but it provides a rich set of
    cryptographic instructions for validating signatures and cryptographic entities. By using these instructions and
    passing signatures as parameters to methods a programmer can implement the required logic for authorizing any
    operation.

    \emph{The Algorand virtual machine has no instructions for issuing cryptographic signatures.}

    In addition to signatures, a method can verify its invoker by using \texttt{get\_parent} instruction. This
    instruction gets the \texttt{applicationID} of the smart contract that is one level deeper than the current
    smart contract in the call stack. In other words, it gives the smart contract that has invoked the current smart
    contract. (if any)

    \subsubsection{Heap Allocation Instructions}

    ...

    \chapter{Implementation}\label{ch:implementation}

@@

    To mitigate this problem, for calculating a user's stake at the time step \(t\), instead of using the raw ALGO
    balance, we use the minimum of a \emph{trust value} the system has calculated for the user and the user's
    ALGO balance:
    \[
        Stake_{u,t} = \min (Balance_{u,t}, Trust_{u,t})
    \]
    For estimating the value of \(Trust_{u,t}\) we use the following formula:
    \[
        Trust_{u,t} = \max (M_{u,t}, \beta Balance_{u,t})
    \]
    Where \(M_{u,t}\) is the exponential moving average of the ALGO balance of the user \(u\) at the time step \(t\),
    and \(\beta\) is a constant between \(0\) and \(1\) determining the minimum trust value of a user.

    The agreement protocol, at the time step \(t\), will use \(\sum_{u}Stake_{u,t}\) to determine the required number of
    votes for the confirmation of a block. At the start of the agreement protocol, \(\beta\) is initialized to \(0\).
    If after confirming a block we have:
    \[
        \sum_{u}Stake_u < \gamma \sum_{u} Balance_u
    \]
    The protocol will increase the value of \(\beta\) to make sure that the total stake of the system goes high enough.
    If after confirming a block we have \(\beta > 0\) and:
    \[
        \sum_{u}Stake_u > \lambda \sum_{u} Balance_u
    \]
    The protocol will decrease the value of \(\beta\) until it eventually reaches \(0\) again.

    In our system a user who held ALGOs and participated in the consensus for a long time is more trusted
    than a user with a higher balance whose balance has increased recently. An attacker who has obtained a large
    amount of ALGOs, also needs to hold them for a long period of time before being able to attack our system.

@@

    A smart contract oracle is a full AVM emulator that keeps a full local copy of AVM memory and can emulate AVM
    execution without accessing a ZK-EDB. Smart contract oracles can be used for reporting useful information about
    \texttt{avmCall} transactions such as accessed AVM heap pages or code area blocks, exact amount of execution cost,
    and soon.
